Ricci-Ollivier curvature of two random walks on rooted phylogenetic subtree-prune-regraft graph

### 2015-02-15 curvature.tex
fixed a spurious maximized.
Abstract and first paragraph.

\begin{document}
\title[Curvature of phylogenetic subtree-prune-regraft graph]{Ricci-Ollivier curvature of two random walks on rooted phylogenetic subtree-prune-regraft graph}
\author[Whidden]{Christopher Whidden}
\author[Matsen]{Frederick Matsen}
\address{}
\thanks{}

@@

\begin{abstract}
\end{abstract}

@@

\section{Introduction}

Statistical phylogenetics is.
Likelihood of generating the observed sequences via a continuous time Markov chain.
Likelihood-based phylogenetics gives an optimality criterion that comes from both branch length and topology.
Bayesian versus maximum likelihood (ML).

Phylogenetic algorithms do MCMC or heuristic local search.
In both settings, one needs a means of describing "nearby" trees.
There are a number of such means of doing so.
SPR.
SPR graph.
For MCMC moves, one would like to ensure good mixing.
For ML moves, we are concerned about local optima.
We will call these "graph effects"-- can't get easily from one spot to another because of structure of the graph.
We use "random walks" as being a general notion of movement around a graph, including MCMC.

@@

The relative change in degree, $\frac{\degree{T}}{\degree{S}}$, can also be written as $\frac{\degree{T}}{\degree{T} + (\degree{S} - \degree{T})$}.
By (1), we have that $\degree{S} - \degree{T} \le \frac{1}{2} (n-2)^2 $.
Thus, $\frac{\degree{T}}{\degree{S}} \ge \frac{\degree{T}}{\degree{T} + \frac{1}{2} (n-2)^2} $.
This bound is maximized when $\degree{T}$ is minimized and, again, $\degree{T} \ge 3n^2 - 13n + 14$ by Lemma~\ref{lem:degree_extremes}.
	We then have:
	\begin{align*}
		\frac{\degree{T}}{\degree{S}} &\ge \frac{3n^2 - 13n + 14}{3n^2 - 13n + 14 + \frac{1}{2}(n-2)^2} \\
		&\ge \frac{3n^2 - 13n + 14}{3.5n^2 - 15n + 16}
	\end{align*}
	Thus (2) and (3) follow.

>>>>

\begin{document}
\title[Curvature of phylogenetic subtree-prune-regraft graph]{Ricci-Ollivier curvature of two random walks on rooted phylogenetic subtree-prune-regraft graph}
\author[Whidden]{Christopher Whidden}
\author[Matsen]{Frederick A Matsen IV}
\address{}
\thanks{}

@@

\begin{abstract}
Statistical phylogenetics inference methods use tree rearrangement operations to perform either local hill-climbing search or Markov chain Monte Carlo across tree topologies.
The canonical class of such moves are the subtree-prune-regraft (SPR) moves that remove a subtree and reattach it somewhere else via the cut edge of the subtree.
Phylogenetic trees and such moves naturally form the vertices and edges of a graph, such that tree search algorithms perform a (potentially stochastic) traversal of the graph.
Despite the centrality of such graphs in phylogenetic inference, rather little is known about their large-scale properties.
In this paper we learn about the rooted-tree version of the graph, known as the rSPR graph, by calculating the recently-developed notion of Ricci curvature for pairs of vertices in the rSPR graph with respect to two simple random walks on the rSPR graph.
This notion of curvature is especially relevant to stochastic search on the graph.
By proving theorems and direct calculation, we find a remarkable diversity of different curvatures on the rSPR graph for pairs of vertices separated by the same distance.
This indicates significant structure of the graph, and a greater understanding of this structure could ultimately lead to improved strategies for tree search.

We find that indeed there is
\end{abstract}

@@

\section{Introduction}
Phylogenetic methods, which reconstruct evolutionary trees from DNA or RNA data, are of fundamental importance to modern biology.
Statistical phylogenetics forms the currently most popular means of reconstructing phylogenetic trees, in which the tree is viewed as an unknown parameter in a likelihood-based statistical inference problem.
The likelihood function in this setting is the likelihood of generating the observed sequences via a continuous time Markov chain (CTMC) evolving down the tree starting from a sequence assumed to be sampled from the stationary distribution (cite{Felsenstein}).
The lengths of the branches of the phylogenetic tree give the ``time'' parameter in the CTMC, where the generated sequence accrues mutations, typically in an IID manner across sites.
Thus likelihood-based phylogenetics gives an optimality criterion that comes from both branch length and topology.
Within this framework researchers typically choose either a Bayesian approach, in which a computer program approximates the posterior distribution of trees and their associated parameters, or a maximum likelihood (ML) approach, in which a computer program searches across tree topologies and continuous parameters in order to find the combination with the highest likelihood.

Here we focus on the discrete search part.
Phylogenetic algorithms do MCMC or heuristic local search.
In both settings, one needs a means of describing "nearby" trees.
There are a number of such means of doing so.
SPR.
SPR graph.
For MCMC moves, one would like to ensure good mixing.
For ML moves, we are concerned about local optima.
We will call these "graph effects"-- can't get easily from one spot to another because of structure of the graph.
We use "random walks" as being a general notion of movement around a graph, including MCMC.

@@

The relative change in degree, $\frac{\degree{T}}{\degree{S}}$, can also be written as $\frac{\degree{T}}{\degree{T} + (\degree{S} - \degree{T})$}.
By (1), we have that $\degree{S} - \degree{T} \le \frac{1}{2} (n-2)^2 $.
Thus, $\frac{\degree{T}}{\degree{S}} \ge \frac{\degree{T}}{\degree{T} + \frac{1}{2} (n-2)^2} $.
This bound is minimized when $\degree{T}$ is minimized and, again, $\degree{T} \ge 3n^2 - 13n + 14$ by Lemma~\ref{lem:degree_extremes}.
	We then have:
	\begin{align*}
		\frac{\degree{T}}{\degree{S}} &\ge \frac{3n^2 - 13n + 14}{3n^2 - 13n + 14 + \frac{1}{2}(n-2)^2} \\
		&\ge \frac{3n^2 - 13n + 14}{3.5n^2 - 15n + 16}
	\end{align*}
	Thus (2) and (3) follow.
